CL-PFU: A suite of R packages for energy conversion chain analysis

### 2023-08-08 paper.Rmd
Working on EA's comments.

---
title: 'MR-PFU: A suite of R packages for energy conversion chain analysis'
tags:
- energy
- exergy
- R
- matrix mathematics
- energy efficiency
authors:
- name: Matthew Kuperus Heun
  orcid: 0000-0002-7438-214X
  affiliation: 1
- name: Zeke Marshall
  orcid: 0000-0001-9260-7827
  affiliation: 2
- name: Paul E. Brockway
  orcid: 0000-0001-6925-8040
  affiliation: 2
- name: Emmanuel Aramendia
  orcid: 0000-0002-5964-6776
  affiliation: 2
affiliations:
- name: Engineering Department, Calvin University
  index: 1
- name: School of Earth and Environment, Leeds University
  index: 2
date: "20 July 2023"
bibliography: paper.bib
output:
  html_document:
    keep_md: yes
---

@@

We created a suite of `R` packages 
to assist SEA practitioners to analyze 
energy movement through society. 
The new packages enable analysis of any country in the world
across timespans of decades or longer.
In short, the packages enable, for the first time, scalable SEA.
We used the new packages to create the MR-PFU database,^["MR-PFU" 
is an initialism for "Multi-Region
Primary, Final, Useful."
The MR-PFU database is "multi-regional" in the sense that
it contains many countries as well as 
continental and world aggregations.
Technically speaking, 
we create [matsindf][link-matsindf] data frames, 
not SQL or similar databases.] 
a new resource for the SEA community [@Marshall:2023aa].

@@

Broadly speaking, 
four calculation steps are required to create a MR-PFU database. 
Each step is assisted by functions in one or more new `R` packages.
First, the IEA's primary- and final-stage WEEB data must be converted
to RUVY matrices for each country and year, 
a task accomplished by functions in the [IEATools][link-ieatools] package [@Heun-IEATools:2023]. 
Second, human and animal muscle work must be calculated from 
[International Labor Organization][link-ilo] (ILO) and
[Food and Agriculture Organization][link-fao] (FAO) data,
following the methodology of @Steenwyk:2022ww, 
using functions in the [MWTools][link-mwtools] package [@Marshall:2023ab]. 
Third, the IEA's primary- and final-stage ECC data 
must be extended to the useful stage by
(a) allocating final stage energy to end-use machines and
(b) multiplying allocated final energy by the 
final-to-useful efficiency of each machine.
This task is accomplished by functions in the
[Recca][link-recca] [@Heun-Recca:2023]
and 
[PFUDatabase][link-pfudatabase] [@Heun-PFUDatabase:2023]
packages.
Fourth, ECCs must be converted from energy terms to exergy terms. 
This step is performed by functions in the [Recca][link-recca] package.

@@

Create a MR-PFU database from the GHA and ZAF data contained in the IEATools package.   
Create the folder structure for the MR-PFU database.   
Store all necessary data in this repository.   
Execute the commands to produce the database, 
albeit for only two countries and two years.

@@

The new `R` packages described in this paper
enable the creation of databases of primary, final, and useful
energy and exergy conversion chains.
Such databases will enable advances in the societal exergy analysis community
and lead to greater insights 
about the role of energy 
in economic growth and human well-being.

>>>>

---
title: 'CL-PFU: A suite of R packages for energy conversion chain analysis'
tags:
- energy
- exergy
- R
- matrix mathematics
- energy efficiency
authors:
- name: Matthew Kuperus Heun
  orcid: 0000-0002-7438-214X
  affiliation: 1
- name: Zeke Marshall
  orcid: 0000-0001-9260-7827
  affiliation: 2
- name: Paul E. Brockway
  orcid: 0000-0001-6925-8040
  affiliation: 2
- name: Emmanuel Aramendia
  orcid: 0000-0002-5964-6776
  affiliation: 2
affiliations:
- name: Engineering Department, Calvin University
  index: 1
- name: School of Earth and Environment, Leeds University
  index: 2
date: "20 July 2023"
bibliography: paper.bib
output:
  html_document:
    keep_md: yes
---

@@

We created a suite of `R` packages 
to assist SEA practitioners to analyze 
energy movement through society. 
The new packages enable analysis of any country in the world
across timespans of decades or longer.
In short, the packages enable, for the first time, scalable SEA.
We used the new packages to create the CL-PFU database,^["CL-PFU" 
is an initialism for "Country-Level
Primary, Final, Useful."
The CL-PFU database is "multi-regional" in the sense that
it contains many countries as well as 
continental and world aggregations.
Technically speaking, 
we create [matsindf][link-matsindf] data frames, 
not SQL or similar databases.] 
a new resource for the SEA community [@Marshall:2023aa].

@@

Broadly speaking, 
four calculation steps are required to create a CL-PFU database. 
Each step is assisted by functions in one or more new `R` packages.
First, the IEA's primary- and final-stage WEEB data must be converted
to RUVY matrices for each country and year, 
a task accomplished by functions in the [IEATools][link-ieatools] package [@Heun-IEATools:2023]. 
Second, human and animal muscle work must be calculated from 
[International Labor Organization][link-ilo] (ILO) and
[Food and Agriculture Organization][link-fao] (FAO) data,
following the methodology of @Steenwyk:2022ww, 
using functions in the [MWTools][link-mwtools] package [@Marshall:2023ab]. 
Third, the IEA's primary- and final-stage ECC data 
must be extended to the useful stage by
(a) allocating final stage energy to end-use machines and
(b) multiplying allocated final energy by the 
final-to-useful efficiency of each machine.
This task is accomplished by functions in the
[Recca][link-recca] [@Heun-Recca:2023]
and 
[PFUDatabase][link-pfudatabase] [@Heun-PFUDatabase:2023]
packages.
Fourth, ECCs must be converted from energy terms to exergy terms. 
This step is performed by functions in the [Recca][link-recca] package.

@@

Create a CL-PFU database from the GHA and ZAF data contained in the IEATools package.   
Create the folder structure for the CL-PFU database.   
Store all necessary data in this repository.   
Execute the commands to produce the database, 
albeit for only two countries and two years.

@@

The new `R` packages described in this paper
enable the creation of databases of primary, final, and useful
energy and exergy conversion chains.
Such databases will enable advances in the societal exergy analysis community
and lead to greater insights 
about the role of energy 
in economic growth and human well-being.
An example of early work using the CL-PFU database
is the calculation of useful stage Energy Return On Investment for fossil fuels [a], 
for which the ECCTools [b] and EROITools [c] packages were developed, 
expanding the software capabilities introduced here."

***** Add Emmanuel references *****

[a]: Aramendia E., Brockway P., Taylor P., Norman J., Heun M., Marshall Z. (2023), “Chapter 4: Estimation of fossil fuels useful stage energy return on investment and implications for
renewable energy systems” in Exploring the useful energy implications of the global energy transition: a net energy perspective, PhD thesis, University of Leeds.
[b]: Emmanuel Aramendia and Matthew K. Heun. ECCTools: Tools for modifying the Energy Conversion Chain. R package version 0.1.5. 2022. doi: 10.5281/zenodo.7692105.
[c]: Emmanuel Aramendia and Matthew K. Heun. EROITools: Tools for calculating and aggregating Energy Return On Investment values. Package version: v0.1.1. 2023. doi: 10 . 5281 / zenodo .7696869.
